# Detecting resting behavior of wild boar in GPS data
by Nina Poglic & Mirjam Kurz

MSc ENR, ZHAW
Patterns and Trends in Environmental Data

Date: 3.7.2022


## Introduction


## Material and Methods
### Dataset
The GPS data describing the movement of the wild boar and habitat information was provided by the ZHAW. 
Each fix point contains the following attributes:
•	Animal ID ("TierID")
•	Animal name ("TierName")
•	Datetime ("DatetimeUTC")
•	X- and Y-coordinate 
•	Daytime ("Day")
The additional information on moon illumination and the collar ID were not used for this project.
GPS collars were fitted to 19 wild boar and were programmed to attempt a GPS location fix every 15 minutes. The total study time lasted for 874 days from the 28th of May 2014 to the 18th of October 2016. However, not all animals were monitored at the same time and some individuals show gaps in their sampling interval (see Figure 1). 

## Results



## R code

#### Packages, loading data and pre-processing
```{r}
library(ComputationalMovementAnalysisData)

ComputationalMovementAnalysisData::wildschwein_BE

library(readr)        # to import tabular data (e.g. csv)
library(dplyr)        # to manipulate (tabular) data
library(ggplot2)      # to visualize data
library(sf)           # to handle spatial vector data
library(terra)        # To handle raster data
library(lubridate)    # To handle dates and times
library(zoo)

load(file='wildschwein_BE.rda')

wildschwein_BE <- st_as_sf(wildschwein_BE, coords = c("E", "N"), crs = 2056, remove = FALSE)

crop_fanel <- read_sf("Feldaufnahmen_Fanel.gpkg") 				#Reading in habitat data

wildschwein_BE <- 
  st_join(wildschwein_BE, crop_fanel) #Annotating habitat to all points

wildschwein_BE$day[wildschwein_BE$day != "Tag"] <- "Nacht" 		#Sorting everything into day or night
```

## Steplengths
```{r}
wildschwein_BE <- wildschwein_BE %>%
  group_by(TierID) %>%
  mutate(steplength = sqrt((E- lead(E,1))^2 + (N -lead(N,1))^2)) #Calculating steplength

mean(wildschwein_BE$steplength, na.rm=TRUE) 	#Calculating mean steplength

median(wildschwein_BE$steplength, na.rm=TRUE) 	#Calculating median steplength

all_January2015 <- wildschwein_BE %>%				#Making dataframe for all animals in one week in 2015
  filter(DatetimeUTC >=as.Date("2015-01-01")& DatetimeUTC <=as.Date("2015-01-08"))
  
ggplot(all_January2015, mapping = aes(DatetimeUTC, steplength))+ 	#Visualizing steplength of all_January2015 
  geom_line(aes(colour=TierName))+
  facet_wrap(~TierName)
```
## Stationary segments
```{r}
window_k12 <- wildschwein_BE
window_k12$k12 <- rollmean(wildschwein_BE$steplength, k=12, fill = NA, align = "left") #calculate mean distance for rolling window k12

window_k12 <- window_k12 %>% 
  ungroup() %>%
  mutate(stationary_k12_10 = k12 < 10) 				#Assigning values below 10 as stationary, in new dataframe

rle_id <- function(vec){						#Function for making segments
  x <- rle(vec)$lengths
  as.factor(rep(seq_along(x), times=x)) 
}

window_k12 <-window_k12 %>% 					#Making segments
  mutate(
    segment_ID = rle_id(stationary_k12_10) 
  )


k12_segments <- window_k12 %>%					#Calculating time span of each stationary segment
  st_drop_geometry() %>%
  filter(stationary_k12_10 == "TRUE") %>%
  group_by(TierName, segment_ID) %>%
  summarise(min = min(DatetimeUTC), max = max(DatetimeUTC)) %>%  
  mutate(timediff = as.integer(difftime(max, min, units = "mins")))


segments_E <- window_k12 %>%					#Giving each segment the first coordinate of the segment
  st_drop_geometry %>%
  group_by(segment_ID) %>%
  summarise(segment_E = first(E))

segments_N <- window_k12 %>%
  st_drop_geometry() %>%
  group_by(segment_ID) %>%
  summarise(segment_N = first(N))

k12_segments <-
  left_join(k12_segments, segments_E, by = "segment_ID")

k12_segments <-
  left_join(k12_segments, segments_N, by = "segment_ID")

k12_segments <- st_as_sf(k12_segments, coords = c("segment_E", "segment_N"), crs = 2056, remove = FALSE)

k12_segments_2h <- k12_segments %>% 			#Filtering out all stationary segments below 2h
  filter(timediff > 120)
  
```
## Exploring stationary segments
```{r}
mean(k12_segments_2h$timediff)					#Calculating mean of all stationary segments

median(k12_segments_2h$timediff)				#Calculating median of all stationary segments
	
mean_stationary_animal <- k12_segments_2h %>% 		#Calculating mean stationary time per animal
  st_drop_geometry() %>%
  group_by(TierName) %>%
  summarise_at(vars(timediff), list(name = mean))

write.table(mean_stationary_animal)


k12_segments_2h <- k12_segments_2h %>%    #Adding hour of start and end of a segment
  mutate(hour_min = hour(min))

k12_segments_2h <- k12_segments_2h %>%
  mutate(hour_max = hour(max))

hour_min <- table(k12_segments_2h$hour_min) #Visualiing start of segments
barplot(hour_min)

hour_max <- table(k12_segments_2h$hour_max) #Visualiing end of segments
barplot(hour_max)
```



